---
title: "DIABLO BRCA — 03: DIABLO Multi-Omics Integration"
author: "Multiomics Integration Demo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 6, dpi = 150,
  fig.path = "../../results/diablo/figures/"
)
```

## Overview

DIABLO (Data Integration Analysis for Biomarker discovery using Latent cOmponents)
extends sPLS-DA to multiple data blocks simultaneously. Its objective function
jointly maximises:
1. **Discrimination** between PAM50 subtypes (sPLS-DA objective per block)
2. **Cross-block correlation** (controlled by the design matrix)

This produces components that are simultaneously discriminant AND capture
coordinated biology across omics layers.

**Key hyperparameters**:
- **Design matrix**: Off-diagonal values (0–1) trade off discrimination vs. cross-block correlation
- **ncomp**: Number of multi-block components (tuned by CV)
- **keepX**: Features retained per block per component (tuned by CV)

---

## Setup

```{r libraries}
library(mixOmics)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)

theme_set(theme_bw(base_size = 12))

subtype_colors <- c(Basal = "#E41A1C", Her2 = "#FF7F00", LumA = "#4DAF4A")

saved     <- readRDS("../../results/diablo/breast_TCGA_processed.RDS")
X_train   <- saved$X_train
Y_train   <- saved$Y_train
X_test    <- saved$X_test
Y_test    <- saved$Y_test

# Load single-omics baseline BERs for comparison
baseline_res <- readRDS("../../results/diablo/single_omics_results.RDS")
baseline_ber  <- setNames(
  baseline_res$baselines |> map_dbl(~ .x$min_ber),
  baseline_res$baselines |> map_chr(~ .x$block)
)

cat("Data and baselines loaded.\n")
```

---

## 1. Design Matrix

```{r design-matrix}
# The design matrix A (n_blocks × n_blocks) controls how strongly DIABLO
# enforces correlation between each pair of blocks during optimisation.
#
# Design = 0.1 (recommended default):
#   Prioritises discrimination (the supervised goal) while applying mild
#   cross-block regularisation. This improves stability and ensures selected
#   features from different blocks are biologically co-regulated, not just
#   accidentally co-selected.
#
# Design = 1: full correlation — the model becomes more like multi-block CCA,
#   which can reduce discriminative power if blocks are not strongly correlated.
# Design = 0: pure discriminant analysis per block — no cross-block coordination.
#
# Biological motivation for 0.1: mRNA, miRNA, and proteomics capture different
# regulatory layers. Their natural cross-block correlation is moderate (as seen
# in Notebook 1 PCA). A value of 0.1 acknowledges this moderate coupling
# without dominating the supervised objective.

n_blocks    <- length(X_train)
block_names <- names(X_train)

design <- matrix(0.1, nrow = n_blocks, ncol = n_blocks,
                 dimnames = list(block_names, block_names))
diag(design) <- 0

cat("Design matrix:\n")
print(design)
```

---

## 2. Initial DIABLO Model (Untuned) for ncomp Selection

```{r initial-model}
# Train a non-sparse DIABLO model with 5 components to estimate
# the optimal number of components via CV. Using all features here
# gives a reliable ncomp estimate before introducing sparsity.

diablo_init <- block.splsda(
  X      = X_train,
  Y      = Y_train,
  ncomp  = 5,
  design = design
)

cat("Initial (untuned) DIABLO model trained.\n")
```

---

## 3. Tune Number of Components

```{r tune-ncomp, eval=TRUE}
# perf() runs stratified M-fold CV on the DIABLO model.
# We use 10 folds and 10 repeats for robust error estimates.
# BER (Balanced Error Rate) is our optimisation metric.

FORCE_TUNE  <- FALSE
ncomp_path  <- "../../results/diablo/perf_ncomp.RDS"

if (!file.exists(ncomp_path) || FORCE_TUNE) {
  set.seed(42)
  perf_init <- perf(
    diablo_init,
    validation  = "Mfold",
    folds       = 10,
    nrepeat     = 10,
    progressBar = FALSE
  )
  saveRDS(perf_init, ncomp_path)
  cat("ncomp tuning complete.\n")
} else {
  perf_init <- readRDS(ncomp_path)
  cat("Pre-computed ncomp tuning loaded.\n")
}

plot(perf_init)
```

```{r choose-ncomp}
# Select ncomp at minimum BER (mahalanobis distance is most stable)
ber_weighted <- perf_init$WeightedVote.error.rate

if (!is.null(ber_weighted)) {
  ber_comp    <- ber_weighted[, "BER", "mahalanobis.dist"]
  optimal_ncomp <- which.min(ber_comp)
  cat("Optimal ncomp:", optimal_ncomp,
      "| Min BER:", round(min(ber_comp) * 100, 1), "%\n")
} else {
  # Fallback for older mixOmics versions
  optimal_ncomp <- 2
  cat("Using default ncomp = 2 (3-class minimum)\n")
}
```

---

## 4. Tune keepX: Sparse Feature Selection

```{r tune-keepx, eval=TRUE}
# tune.block.splsda() searches a grid of keepX values per block.
# It selects the combination minimising BER in cross-validation.
# Smaller keepX → sparser, more interpretable, potentially lower
# test accuracy; larger keepX → more features but risk overfitting.

keepX_path <- "../../results/diablo/tune_keepX.RDS"

if (!file.exists(keepX_path) || FORCE_TUNE) {

  # Grid of feature counts to try per block
  test_keepX <- list(
    mRNA       = c(5, 10, 15, 20, 25, 30),
    miRNA      = c(5, 10, 15, 20),
    proteomics = c(5, 10, 15, 20)
  )

  set.seed(42)
  tune_keepX <- tune.block.splsda(
    X           = X_train,
    Y           = Y_train,
    ncomp       = optimal_ncomp,
    test.keepX  = test_keepX,
    design      = design,
    validation  = "Mfold",
    folds       = 10,
    nrepeat     = 5,
    dist        = "mahalanobis.dist",
    progressBar = FALSE
  )

  saveRDS(tune_keepX, keepX_path)
  cat("keepX tuning complete.\n")
} else {
  tune_keepX <- readRDS(keepX_path)
  cat("Pre-computed keepX tuning loaded.\n")
}

optimal_keepX <- tune_keepX$choice.keepX
cat("Optimal keepX per block:\n")
print(optimal_keepX)
```

---

## 5. Train Final DIABLO Model

```{r final-model}
diablo_final <- block.splsda(
  X      = X_train,
  Y      = Y_train,
  ncomp  = optimal_ncomp,
  keepX  = optimal_keepX,
  design = design
)

cat("Final DIABLO model trained.\n")
cat("  Components:", optimal_ncomp, "\n")
walk2(names(optimal_keepX), optimal_keepX, function(b, k)
  cat(sprintf("  %s: keepX = %s\n", b, paste(k, collapse=", ")))
)
```

---

## 6. LOO and M-Fold Cross-Validation

```{r cv-final, eval=TRUE}
# Cross-validate the final sparse DIABLO model for honest performance estimates.
# We run both 10-fold CV (faster) and optionally LOO (leave-one-out, exact but slow).

cv_path <- "../../results/diablo/cv_final.RDS"

if (!file.exists(cv_path) || FORCE_TUNE) {
  set.seed(42)
  cv_final <- perf(
    diablo_final,
    validation  = "Mfold",
    folds       = 10,
    nrepeat     = 10,
    progressBar = FALSE
  )
  saveRDS(cv_final, cv_path)
} else {
  cv_final <- readRDS(cv_path)
}

plot(cv_final, col = color.mixo(1:3))
```

```{r loo-cv, eval=FALSE}
# LOO cross-validation: most unbiased estimate for small samples,
# but computationally expensive (n_samples × nrepeat iterations).
# Set eval=TRUE for final publication-quality results.

set.seed(42)
cv_loo <- perf(
  diablo_final,
  validation  = "loo",
  progressBar = TRUE
)

cat("LOO BER:\n")
print(cv_loo$WeightedVote.error.rate[, "BER", "mahalanobis.dist"])
```

```{r cv-results}
# Extract and display final CV results
if (!is.null(cv_final$WeightedVote.error.rate)) {
  cv_ber <- cv_final$WeightedVote.error.rate[optimal_ncomp, "BER", "mahalanobis.dist"]
  cat(sprintf("DIABLO 10-fold CV BER (weighted vote, comp %d): %.1f%%\n",
              optimal_ncomp, 100 * cv_ber))

  cat("\nSingle-omics BER baseline:\n")
  walk2(names(baseline_ber), baseline_ber, function(b, ber)
    cat(sprintf("  %s: %.1f%%\n", b, 100 * ber))
  )
  cat(sprintf("\nImprovement over best single-omics: %.1f percentage points\n",
              100 * (min(baseline_ber) - cv_ber)))
}
```

---

## 7. Sample Score Plots

```{r plotIndiv-all, fig.height=6}
# Plot all four panels: one per block + weighted average consensus.
# The weighted average shows the integrated sample representation.

plotIndiv(
  diablo_final,
  comp          = c(1, 2),
  blocks        = c("mRNA", "miRNA", "proteomics", "weighted.average"),
  ind.names     = FALSE,
  legend        = TRUE,
  col.per.group  = subtype_colors,
  ellipse       = TRUE,
  ellipse.level = 0.95,
  title         = "DIABLO component scores — all blocks + consensus"
)
```

```{r arrow-plot, fig.height=5}
# Arrow plot: tail = mRNA position, head = miRNA position per sample.
# Short arrows = both views agree on subtype positioning → strong consensus.
# Long arrows = discordance between mRNA and miRNA representations.

plotArrow(
  diablo_final,
  ind.names     = FALSE,
  group         = Y_train,
  col.per.group  = subtype_colors,
  title         = "Arrow plot: mRNA (tail) → miRNA (head) per sample"
)
```

---

## 8. Test Set Evaluation

```{r test-evaluation}
# Predict on the held-out test set — never used during training or tuning.
test_pred <- predict(
  diablo_final,
  newdata = X_test,
  dist    = "mahalanobis.dist"
)

# Weighted vote predictions from the final component
pred_labels <- test_pred$WeightedVote$mahalanobis.dist[, optimal_ncomp]

# Confusion matrix
conf_mat    <- table(Predicted = pred_labels, Actual = Y_test)
overall_acc <- sum(diag(conf_mat)) / sum(conf_mat)

cat("=== DIABLO Test Set Results ===\n\n")
cat("Confusion matrix:\n")
print(conf_mat)
cat(sprintf("\nOverall accuracy: %.1f%%\n", 100 * overall_acc))
cat(sprintf("Overall BER:      %.1f%%\n",
    100 * (1 - mean(diag(conf_mat) / colSums(conf_mat)))))
```

```{r confusion-heatmap, fig.height=4}
as.data.frame(conf_mat) |>
  ggplot(aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(colour = "white") +
  geom_text(aes(label = Freq), size = 5, fontface = "bold") +
  scale_fill_gradient(low = "white", high = "#2c7bb6") +
  scale_x_discrete(position = "top") +
  labs(
    title    = "DIABLO test set confusion matrix",
    subtitle = sprintf("Overall accuracy: %.1f%% | Multi-omics weighted vote",
                       100 * overall_acc),
    x = "True subtype", y = "Predicted subtype", fill = "Count"
  )
```

```{r accuracy-comparison, fig.height=4}
# Compare DIABLO vs single-omics test accuracy
single_omics_acc <- baseline_res$test_results |>
  select(block, overall_acc)

bind_rows(
  single_omics_acc,
  data.frame(block = "DIABLO (mRNA + miRNA, weighted vote)", overall_acc = overall_acc)
) |>
  ggplot(aes(x = reorder(block, overall_acc), y = overall_acc,
             fill = block == "DIABLO (mRNA + miRNA, weighted vote)")) +
  geom_col(show.legend = FALSE, alpha = 0.85) +
  geom_text(aes(label = sprintf("%.1f%%", 100 * overall_acc)),
            hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent, limits = c(0, 1.15)) +
  scale_fill_manual(values = c("FALSE" = "#4393c3", "TRUE" = "#d73027")) +
  labs(
    title    = "Test accuracy: DIABLO vs single-omics baselines",
    subtitle = "DIABLO (red) uses available test blocks (mRNA + miRNA)",
    x = NULL, y = "Accuracy (test set)"
  )
```

---

## 9. Save Final Model

```{r save-model}
saveRDS(
  list(
    model         = diablo_final,
    cv            = cv_final,
    test_pred     = test_pred,
    conf_mat      = conf_mat,
    overall_acc   = overall_acc,
    optimal_ncomp = optimal_ncomp,
    optimal_keepX = optimal_keepX,
    design        = design
  ),
  "../../results/diablo/DIABLO_final_model.RDS"
)
cat("Final model saved.\n")
```

---

## Results Summary

| Model | CV BER (10-fold) | Test Accuracy |
|-------|-----------------|--------------|
| mRNA only | `r sprintf("%.1f%%", 100*baseline_ber["mRNA"])` | `r sprintf("%.1f%%", 100*baseline_res$test_results$overall_acc[baseline_res$test_results$block=="mRNA"])` |
| miRNA only | `r sprintf("%.1f%%", 100*baseline_ber["miRNA"])` | `r sprintf("%.1f%%", 100*baseline_res$test_results$overall_acc[baseline_res$test_results$block=="miRNA"])` |
| proteomics only | `r sprintf("%.1f%%", 100*baseline_ber["proteomics"])` | `r sprintf("%.1f%%", 100*baseline_res$test_results$overall_acc[baseline_res$test_results$block=="proteomics"])` |
| **DIABLO (mRNA + miRNA weighted vote on test set)** | **`r if(exists('cv_ber')) sprintf("%.1f%%", 100*cv_ber) else "—"`** | **`r sprintf("%.1f%%", 100*overall_acc)`** |

**Key conclusion**: In this dataset split, test-time DIABLO evaluation uses the available
blocks (mRNA + miRNA) and outperforms either single-omics baseline. This supports the
value of integrating complementary cross-omics signal.

**→ Next: `04_biomarker_interpretation.Rmd`**
