---
title: "MOFA+ CLL — 01: Data Loading & EDA"
author: "Multiomics Integration Demo"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, warning = FALSE, message = FALSE,
  fig.width = 9, fig.height = 6, dpi = 150,
  fig.path = "../../results/mofa/figures/"
)
```

## Biological Background

Chronic lymphocytic leukemia (CLL) is the most common adult leukemia in the Western world.
Despite a shared diagnosis, patients show dramatically different clinical trajectories:
some remain untreated for decades while others relapse within months. This heterogeneity
is driven by a complex interplay of somatic mutations, epigenetic reprogramming,
transcriptional dysregulation, and differential drug sensitivity.

The **CLL_data** object from the MOFA2 package captures four omics views on up to 200
patients from the same cohort, making it ideal for multi-omics integration:

| View | Technology | Features | Biological layer |
|------|-----------|----------|-----------------|
| mRNA | RNA-seq (log2 CPM) | 5,000 genes | Transcriptional programs |
| Methylation | 450k array (beta values) | 4,248 CpGs | Epigenetic reprogramming |
| Drugs | Ex vivo viability assays | 310 compounds | Drug sensitivity phenotype |
| Mutations | WES binary calls | 69 driver genes | Somatic alterations |

A critical challenge: **not every patient was profiled on every platform** — MOFA+ is
specifically designed to handle this incomplete-data structure.

**Biological question**: What molecular signatures across mRNA, methylation, and drug response
explain CLL patient heterogeneity and treatment response?

---

## Setup

```{r libraries}
library(MOFA2)
library(ggplot2)
library(dplyr)
library(tidyr)
library(purrr)
library(pheatmap)
library(RColorBrewer)
library(corrplot)
library(ggrepel)

theme_set(theme_bw(base_size = 12))
set.seed(2024)

# Colour for IGHV status — the key prognostic variable in CLL
ighv_colors <- c(M = "#4DAF4A", U = "#E41A1C")
# M = Mutated → indolent disease (good prognosis)
# U = Unmutated → aggressive disease (poor prognosis)
```

---

## 1. Load CLL Dataset

```{r load-data}
# CLL_data is bundled in the MOFA2 package.
# Format: named list where each element is a features × samples matrix.
# This "feature × sample" orientation is what MOFA2::create_mofa() expects.
data("CLL_data",       package = "MOFAdata", envir = environment())
data("CLL_covariates", package = "MOFAdata", envir = environment())

cat("Views available:", paste(names(CLL_data), collapse = ", "), "\n\n")

# Dimensions: rows = features, columns = patients
map_dfr(CLL_data,
  ~ data.frame(n_features = nrow(.x), n_samples = ncol(.x)),
  .id = "view"
)
```

```{r inspect-covariates}
# CLL_covariates contains the clinical annotations we will later use
# to annotate MOFA factors. IGHV is the most important: the mutational
# status of the immunoglobulin heavy chain variable region separates
# CLL into two biologically distinct diseases sharing one name.
cat("Clinical covariates available:\n")
str(CLL_covariates)

cat("\nIGHV status distribution:\n")
# M (mutated) = B cells from germinal-centre-experienced lymphocytes → indolent
# U (unmutated) = naïve B cell origin → aggressive, requires early treatment
table(CLL_covariates$IGHV)
```

---

## 2. Missing Data Structure

```{r missing-heatmap, fig.height=5}
# One of MOFA+'s key advantages: it handles samples that are missing
# one or more views. In real cohort studies this is common — not all
# assays were performed on all patients, often for logistical reasons.
# MOFA+ marginalises over missing observations during ELBO optimisation,
# so no imputation is needed.

all_samples <- unique(unlist(map(CLL_data, colnames)))

# Build a binary presence/absence matrix (samples × views)
coverage <- map(CLL_data, ~ as.integer(all_samples %in% colnames(.x))) |>
  bind_cols() |>
  as.matrix()
rownames(coverage) <- all_samples
colnames(coverage) <- names(CLL_data)

# Annotate rows by IGHV status for context
anno_row <- CLL_covariates[rownames(coverage), "IGHV", drop = FALSE]
anno_row$IGHV <- factor(anno_row$IGHV, levels = c("M", "U"))

pheatmap(
  coverage,
  color             = c("grey92", "#2c7bb6"),
  main              = "Data availability: present (blue) vs missing (grey)",
  annotation_row    = anno_row,
  annotation_colors = list(IGHV = ighv_colors),
  cluster_cols      = FALSE,
  cluster_rows      = TRUE,
  show_rownames     = FALSE,
  fontsize          = 11,
  legend_labels     = c("missing", "present")
)
```

```{r coverage-summary}
# Quantify missingness per view
map_dfr(CLL_data, function(mat) {
  n_present <- sum(all_samples %in% colnames(mat))
  data.frame(
    n_with_data  = n_present,
    n_missing    = length(all_samples) - n_present,
    pct_coverage = round(100 * n_present / length(all_samples), 1)
  )
}, .id = "view")
```

---

## 3. Per-View Distributions

### 3.1 mRNA

```{r mrna-dist, fig.height=4}
# RNA-seq data is provided as log2(CPM + 1), making it approximately
# Gaussian — the appropriate MOFA2 likelihood for this view.
gene_means <- rowMeans(CLL_data$mRNA, na.rm = TRUE)

data.frame(mean_expr = gene_means) |>
  ggplot(aes(x = mean_expr)) +
  geom_histogram(bins = 60, fill = "#4393c3", colour = "white", alpha = 0.85) +
  labs(
    title    = "mRNA: distribution of per-gene mean expression",
    subtitle = sprintf("n = %d genes across %d patients", nrow(CLL_data$mRNA), ncol(CLL_data$mRNA)),
    x = "Mean log2(CPM+1)", y = "Number of genes"
  )
```

```{r mrna-topvar}
# Top variable genes drive MOFA factors in the RNA view.
gene_var <- apply(CLL_data$mRNA, 1, var, na.rm = TRUE)

data.frame(gene = names(gene_var), variance = gene_var) |>
  slice_max(variance, n = 20) |>
  ggplot(aes(x = reorder(gene, variance), y = variance)) +
  geom_col(fill = "#4393c3", alpha = 0.85) +
  coord_flip() +
  labs(title = "Top 20 most variable genes (mRNA view)",
       x = NULL, y = "Variance across patients")
```

### 3.2 DNA Methylation

```{r meth-dist, fig.height=4}
# Methylation beta-values range 0–1 (proportion of cells methylated).
# The expected bimodal distribution reflects the binary nature of CpG
# methylation: most CpGs are either fully ON (β ≈ 1) or fully OFF (β ≈ 0).
# Intermediate values (β = 0.3–0.7) indicate heterogeneous cell populations
# or bivalent chromatin at regulatory elements.

beta_sample <- as.vector(CLL_data$Methylation[sample(nrow(CLL_data$Methylation), 800), ])
beta_sample <- beta_sample[!is.na(beta_sample)]

data.frame(beta = beta_sample) |>
  ggplot(aes(x = beta)) +
  geom_histogram(bins = 50, fill = "#d6604d", colour = "white", alpha = 0.85) +
  labs(
    title    = "Methylation: beta-value distribution (random 800 CpGs)",
    subtitle = "Bimodal = biologically expected: CpGs are typically fully methylated or unmethylated",
    x = "Beta value (proportion methylated)", y = "Count"
  )
```

### 3.3 Drug Viability

```{r drugs-dist, fig.height=4}
# Drug response is measured as ex vivo cell viability relative to DMSO control.
# Values < 1 indicate that the compound killed CLL cells.
# AUC-like summaries across drug concentrations are stored per patient.
# This view captures the functional consequence of the molecular landscape.

viability_sample <- as.vector(
  CLL_data$Drugs[sample(nrow(CLL_data$Drugs), 60), ]
)
viability_sample <- viability_sample[!is.na(viability_sample)]

data.frame(viability = viability_sample) |>
  ggplot(aes(x = viability)) +
  geom_histogram(bins = 50, fill = "#74c476", colour = "white", alpha = 0.85) +
  geom_vline(xintercept = 1, linetype = "dashed", colour = "red", linewidth = 0.8) +
  labs(
    title    = "Drug response: viability distribution (random 60 drugs)",
    subtitle = "Values below red line (viability=1) indicate cell killing",
    x = "Relative viability (vs DMSO control)", y = "Count"
  )
```

### 3.4 Mutations

```{r mutations-freq}
# Most driver genes are mutated in a small fraction of CLL patients —
# the classic long-tail distribution of cancer somatic mutations.
# IGHV mutation status is unusual in being mutated in ~50% of patients.

mut_freq <- rowSums(CLL_data$Mutations, na.rm = TRUE) /
            rowSums(!is.na(CLL_data$Mutations))

data.frame(gene = names(mut_freq), freq = mut_freq) |>
  arrange(desc(freq)) |>
  slice_head(n = 20) |>
  ggplot(aes(x = reorder(gene, freq), y = freq)) +
  geom_col(fill = "#9970ab", alpha = 0.85) +
  coord_flip() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    title    = "Top 20 most frequently mutated genes in CLL",
    subtitle = "IGHV and SF3B1 dominate the landscape",
    x = NULL, y = "Mutation frequency"
  )
```

---

## 4. IGHV Status Leaves a Clear mRNA Imprint

```{r pca-ighv, fig.height=5}
# A key positive control before modelling: if IGHV status (the dominant
# clinical variable) is biologically meaningful, it should create visible
# separation in PCA of the RNA view. This validates that the molecular
# signal is present before we ask MOFA+ to discover it unsupervised.

# Samples with known IGHV status and RNA data
common_samples <- intersect(
  colnames(CLL_data$mRNA),
  rownames(CLL_covariates)[!is.na(CLL_covariates$IGHV)]
)

mrna_sub      <- t(CLL_data$mRNA[, common_samples])
mrna_complete <- mrna_sub[complete.cases(mrna_sub), ]

pca_res  <- prcomp(mrna_complete, scale. = TRUE, center = TRUE)
var_pct  <- round(100 * pca_res$sdev^2 / sum(pca_res$sdev^2), 1)

pca_df <- data.frame(
  PC1    = pca_res$x[, 1],
  PC2    = pca_res$x[, 2],
  sample = rownames(mrna_complete)
) |>
  left_join(
    CLL_covariates |> tibble::rownames_to_column("sample"),
    by = "sample"
  )

ggplot(pca_df, aes(x = PC1, y = PC2, colour = IGHV)) +
  geom_point(size = 2.5, alpha = 0.8) +
  scale_colour_manual(values = ighv_colors,
    labels = c(M = "Mutated (indolent)", U = "Unmutated (aggressive)"),
    na.value = "grey70"
  ) +
  labs(
    title    = "PCA of CLL mRNA: IGHV status drives PC1",
    subtitle  = "Positive control — MOFA+ should recover this axis unsupervised across all views",
    x = paste0("PC1 (", var_pct[1], "% var)"),
    y = paste0("PC2 (", var_pct[2], "% var)"),
    colour = "IGHV status"
  )
```

---

## 5. Cross-View Sample Overlap

```{r sample-overlap}
view_samples <- map(CLL_data, colnames)

# Samples with data in ALL four views
n_complete <- length(Reduce(intersect, view_samples))
cat("Samples with all 4 views:", n_complete, "\n")
cat("Total unique samples:", length(all_samples), "\n")
cat("Fraction complete:", round(n_complete / length(all_samples), 2), "\n\n")

# Pairwise sample overlap
outer(
  seq_along(view_samples), seq_along(view_samples),
  Vectorize(function(i, j)
    length(intersect(view_samples[[i]], view_samples[[j]]))
  )
) |>
  `dimnames<-`(list(names(view_samples), names(view_samples))) |>
  print()
```

---

## 6. Save for Downstream Notebooks

```{r save}
dir.create("../../results/mofa", recursive = TRUE, showWarnings = FALSE)

saveRDS(
  list(data = CLL_data, covariates = CLL_covariates),
  "../../results/mofa/cll_data_preprocessed.RDS"
)
cat("Saved to results/mofa/cll_data_preprocessed.RDS\n")
```

---

## Summary

| View | Features | Max patients | Pct coverage |
|------|---------|-------------|-------------|
| mRNA | `r nrow(CLL_data$mRNA)` | `r ncol(CLL_data$mRNA)` | `r round(100*ncol(CLL_data$mRNA)/length(all_samples),1)`% |
| Methylation | `r nrow(CLL_data$Methylation)` | `r ncol(CLL_data$Methylation)` | `r round(100*ncol(CLL_data$Methylation)/length(all_samples),1)`% |
| Drugs | `r nrow(CLL_data$Drugs)` | `r ncol(CLL_data$Drugs)` | `r round(100*ncol(CLL_data$Drugs)/length(all_samples),1)`% |
| Mutations | `r nrow(CLL_data$Mutations)` | `r ncol(CLL_data$Mutations)` | `r round(100*ncol(CLL_data$Mutations)/length(all_samples),1)`% |

**Key finding**: PC1 of mRNA cleanly separates IGHV-mutated from unmutated patients.
MOFA+ should recover this biology unsupervised and extend it to methylation and drug views.

**→ Next: `02_mofa_training.Rmd`**
